---
title: "BrightBytes Data Challenge"
output:
  html_notebook: default
  html_document: default
---

###Intro
**Objective**: Use this dataset to help understand what factors are the best predictors of postsecondary enrollment, and offer suggestions to operationalize those insights to help Brightbytes.

**Interpretation**: which factors are the best predictors of gaining a further education after highschool. This includes any additional including college/university and associates degreed. 

**Methodology**: Make the target binary. 1 = Post secondary (anything above highschool), 0 = non-post secondary. Identify factors that are associated with higher rates of post secondary education.


Load libraries and data
```{r}
library(tidyverse)
df <- read_csv('/home/will/link_to_wjburton/Documents/Professional Info/Company Code Challenges/BrightBytes/HSLS_2009_v3_0/students.csv')
```

Based on Data dictionary, drop columns that will not be used, and rename columns to be more interpretable.
```{r}

```

